% PSQF 4143: Section 2
% Brandon LeBeau

# Summarizing Data
- Data in its raw form is often too complex to summarize and understand quickly.
- As such, summarizing the data with descriptive statistics (section 3) or with tables and graphs (this section) can be very helpful.

# Motivating Example
```{r summarize}
# Function to wrap printing.
wrap_print <- function(table, num_row = NULL, num_col = 2) {
  if(is.null(num_row)) {
    num_row <- nrow(table)/2
  }
  ngroups <- nrow(table)/num_row
  
  tmp <- lapply(1:ngroups, function(xx) 
    table[((xx-1)*num_row+1):((xx-1)*num_row + num_row), ])
  tmp <- do.call("cbind", tmp)
  return(tmp)
}

library(ggplot2)
library(dplyr)
title_movie <- c('Mean Girls', 'Notebook, The', 'Troy',
                 'Incredibles, The', 
                 'Harry Potter and the Prisoner of Azkaban',
                 'Napoleon Dynamite', 
                 'Eternal Sunshine of the Spotless Mind',
                 'Crash', '13 Going On 30', 
                 'Girl Next Door, The', 
                 'Dodgeball: A True Underdog Story',
                 'Anchorman: The Legend of Ron Burgundy',
                 'Million Dollar Baby',
                 'Shaun of the Dead', 'White Chicks',
                 'Spider-Man 2', 'Saw', 'Dawn of the Dead',
                 'Van Helsing', 'Cinderella Story, A', 
                 'King Arthur', 'Phantom of the Opera, The',
                 '50 First Dates', 'Aviator, The', 
                 'Kill Bill: Vol. 2', 'Punisher, The',
                 'Village, The', 'Collateral',
                 'Butterfly Effect, The', 'Hellboy')
tmp <- movies %>%
  filter(title %in% title_movie & year == 2004) %>%
  select(title, length)
wrap_print(tmp, num_row = 15)
```

# Ungrouped Frequency Table


# Ungrouped Frequency Table Creation Steps
- To create an ungrouped frequency table:
    1. List all numbers from the minimum to maximum
    2. Count the number that fall within each number
    3. Numbers can have 0 frequency
    
- Strengths:
    - Able to see all the numbers in the distribution
    - Can quickly glance at the range
    - Can quickly see which number is most frequent.
- Weaknesses:
    - Tables can be large depending on range
    - Lose some information, namely who had which value
    
# Grouped Frequency Table


# Grouped Frequency Table Creation Steps
- To create a grouped frequency table:
    1. First need to find the class interval size
       $$ i = Real Upper Limit - Real Lower Limit $$
    2. List out all the class intervals
    3. Count the number that fall within each interval
    
- Strengths:
    - Can quickly see where most values fall
    - 
- Weaknesses:
    - No longer know exact values
    - The choice of class interval size can influence the table

# Additional Notes for Grouped Frequency Tables
- The class intervals should be mutually exclusive
- For quantitative values, no gaps between class intervals
- All class intervals should be the same size
- On average, 10 to 20 class intervals are most appropriate
- Common class interval sizes include: 1, 2, 3, 5, 10, 15, 20, 25, ...
- The lower limit should equal the class size times a constant integer (i.e. 5 * 10 = 50)

# Relative Frequency Distributions
- These are frequency distributions with proportion or percentage of each class interval
- Useful for comparisons of two or more groups especially when the number in each group differ

# Relative Frequency by Groups
```{r freqgroups}
tmp <- movies %>%
  filter(year == 2004 & mpaa %in% c('PG', 'R')) %>%
  select(length, mpaa)
breaks <- seq(10, 180, by = 15)
tmp <- split(tmp, tmp$mpaa)
tmp <- lapply(1:length(tmp), function(xx)
  cut(tmp[[xx]]$length, breaks, right = FALSE))
tmp2 <- data.frame(length = c(tmp[[3]], tmp[[5]]), 
                   mpaa = rep(c('PG', 'R'), 
                              times = c(length(tmp[[3]]),
                                        length(tmp[[5]]))))
tmp_tab <- table(tmp2$length, tmp2$mpaa)
```

